---
title: "Downloading zip files in R"
author: "Tham, Wei Yang"
date: "2017-05-01"
output: html_document
tags:
  - r 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In theory this should be a straightforward task with the `download.file` function, but as I recently found, it's a little trickier than that (but really not that much trickier).[^1] The main things you have to do are 

[^1]: I relied on this [Stack Overflow post](http://stackoverflow.com/questions/3053833/using-r-to-download-zipped-data-file-extract-and-import-data) to figure this out, and it has a few variants on the solution that I'm using here. 

1. Specify `mode = "wb"` in `download.file`
2. Use `unz` to unzip

Example code below. I used a real url in the example that downloads data on NIH grants from [this page](https://exporter.nih.gov/ExPORTER_Catalog.aspx). 

```{r, eval = FALSE}
# download zipped folder
download.file(url = "https://exporter.nih.gov/CSVs/final/RePORTER_PRJ_C_FY2013.zip", 
              destfile = "destination/folder/RePORTER_PRJ_C_FY2013.zip", 
              mode = "wb")

# extract file from zipped folder
x = readr::read_csv(unz(description = "/destination/folder/RePORTER_PRJ_C_FY2013.zip", 
                        filename = "RePORTER_PRJ_C_FY2013.csv"))

```

This is what you should get at the end:

```{r, echo = FALSE, message = FALSE}
readr::read_csv(unz("/Users/weiyangtham/Downloads/g2013.zip", "RePORTER_PRJ_C_FY2013.csv"),
             n_max = 100)
```


