---
title: "Stata's `reshape` in R"
author: "Tham, Wei Yang"
date: "2018-01-13"
output: html_document
tags:
  - r
  - stata
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = FALSE, warning=FALSE}

library(tidyverse)

scores <- data.frame(
id = LETTERS[1:5],
age2000 = 11:15,
age2010 = 21:25,
scores2000 = 96:100,
scores2010 = 100:96)
```

Update: There is an R function called `reshape` from the `stats` package that does the same thing. 

If you've used Stata you might be familiar with its [`reshape`](https://www.stata.com/manuals13/dreshape.pdf) command. `reshape` makes a wide dataset long and vice versa. The equivalent in the `tidyverse` would be the `gather` (wide to long) and `spread` (long to wide) functions from the `tidyr` package. The difference is that `gather` and `spread` work on key-*value* pairs, emphasis on the singular "value", while `reshape` is fine with having multiple values associated with a single key. For example, to reshape the following (fake) wide dataset from this:

```{r, echo = FALSE}
scores %>% knitr::kable()
```

to this:

```{r, echo = FALSE}
twydyverse::gather_multivalue(scores, "year", age2000:scores2010) %>% knitr::kable()
```

In Stata you would do this with something **like**[^1] `reshape long age scores, i(id) j(year)`.

[^1]: I didn't check this code in Stata so could be wrong!

With the `tidyr` functions, you need to first `gather`:

```{r}
scores_vlong = scores %>% tidyr::gather("key2", "value", c(age2000:scores2010))
scores_vlong
```

Then `extract`[^2] and `spread`:

```{r}

scores_vlong %>% 
  tidyr::extract("key2", c("colname", "year"), 
                 regex = "([a-z]+)(\\d+)") %>%
  tidyr::spread("colname", "value")

```

[^2]: The first part of the regex, `"([a-z]+)"`, extracts the word, then `(\\d+)` extracts the digits. 

I've been thinking about writing a function to automate this process for a while and finally got some impetus to do so when [Paul Goldsmith-Pinkham](https://twitter.com/paulgp) asked about the issue on Twitter. Plus it was a good way to practice working with [quasiquotation](http://dplyr.tidyverse.org/articles/programming.html).

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">For <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> users -- how do I reshape wide / long using gather and spread with multiple variables?</p>&mdash; Paul G-P (@paulgp) <a href="https://twitter.com/paulgp/status/952178749075804160?ref_src=twsrc%5Etfw">January 13, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


[`gather_multivalue` and `spread_multivalue`](https://github.com/weiyangtham/twydyverse) are basically wrappers around this sequence of steps.[^3] So now you only need one line:

[^3]: Which in turn are from [Hadley Wickham's Stackoverflow answer](https://stackoverflow.com/questions/25925556/gather-multiple-sets-of-columns)


```{r, eval = FALSE}

# equivalent

gather_multivalue(scores, "year", age2000:scores2010)
gather_multivalue(scores, "year", -id)

```

`gather_multivalue` also asks you to specify a regular expression (regex) for how to extract the key and values. The default regex assumes that the columns are of the form `(word)(number)`. I like that regex gives you some flexibility if you get columns with slightly weird or varying patterns:

```{r, echo = FALSE}
scores_dumb <- data.frame(
id = LETTERS[1:5],
age.2000 = 11:15,
age.2010 = 21:25,
scores_2000 = 96:100,
scores_2010 = 100:96)

scores_dumb %>% knitr::kable()
```

The columns have different separators, `.` and `_`.

```{r}
twydyverse::gather_multivalue(scores_dumb, "year", 
                              age.2000:scores_2010, 
                              regex = "([a-z]+)[\\.|_](\\d+)") 
```

Of course the tradeoff is that you need to specify a regex, but for the purposes of working with column names I don't imagine that's likely to get too complicated.

These functions are available in my personal package:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("weiyangtham/twydyverse")
```


